\chapter{New Commands}
\label{cha:new-commands}

\LaTeX{} also gives you the ability to define your own commands.  This
ability is seldom necessary as a great variety of packages are
distributed with the major \LaTeX{} distributions and many others are
available on the Internet from \LaTeX{}-using organizations and
individuals.  Chances are that most things you want to do are
converged either in \LaTeX{} or an available package.

Should you find yourself in a situation where you need something that
is not available in \LaTeX{} or a package, or you have a bunch of
commands that are used repeatedly, you can define a new command in the
preamble of your document.

\section{Defining New Commands}
\label{sec:defin-new-comm}

The \verb|\newcommand| command allows you to define a command with a
name that is not already in use by \LaTeX{} or any of the packages you
are using with your document.  Usage is as follows:
\begin{verbatim}
\newcommand{name}{definition}
\end{verbatim}

Additionally, you can also add a third argument as follows to specify
the number of arguments for your new command:
\begin{verbatim}
\newcommand{name}[number of arguments]{definition}
\end{verbatim}

The \texttt{name} argument is the name you want for your command.  For
example, \verb|\card| might be the name you want.  The definition
argument is a set of commands that describes what the command should
do.  For example, if you want \verb|\card| to double overline its
argument, you would use \verb|\overline{\overline{}}|.  Thus, you
would add
\begin{verbatim}
\newcommand{\card}{\overline{\overline}}
\end{verbatim}
to the preamble of your document.  Alternatively, you could use
\begin{verbatim}
\newcommand{\card}[1]{\overline{\overline{#1}}}.
\end{verbatim}
This formation is essential if you want to define a command that has
more than one argument.

For example, if you have the following document:
\begin{verbatim}
\documentclass{article}
\newcommand{\card}[1]{\overline{\overline{#1}}}

\begin{document}

\[
\card{A}=\aleph_0
\]

\end{document}
\end{verbatim}
you would get an output that looks like
\[
\card{A}=\aleph_0
\]

\section{Redefining Commands}
\label{sec:redefining-commands}

If a command already exists, you cannot use it as a name for a new
command.  However, if you want to redefine how a command acts, you can
do this with the \verb|\renewcommand| command.  Other than modifying
existing commands, rather than defining new commands,
\verb|\renewcommand| behaves exactly like \verb|\newcommand|.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "reader"
%%% End: 